Thermal reactionome of a temperate ant species

### 2014-05-19 doc/ApTranscriptome_manuscript.Rmd
Edits

We performed a number of comparisons to evaluate the extent to which the thermal reactionomes differed between *A. carolinensis* and *A. picea*. First, we found that the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3), with a similar number of genes with maximum expression at high temperatures, many more genes with maximum expression at intermediate temperatures in *A. carolinensis* and more genes with maximum expression at low temperatures in *A. picea*.

We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate)` TPM greater in *A. carolinensis*. Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$slope)` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r max(t.curvature$p.value, 0.001)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.

@@

By sampling a warmer (*A. carolinensis*) and colder (*A. picea*) climate species, we were able to test if the same genes that are up-regulated in response to thermal stress have increased basal expression in the other species. Partially consistent with this prediction, basal expression was greater in the more southern *A. carolinensis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. carolinensis* basal expression was also greater in *A. carolinensis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`). Moreover, basal expression was also greater in *A. carolinensis* for genes with *Intermediate* expression in either colony. The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A. carolinensis* has greater optimal-temperature expression.

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* and *Bimodal* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity among species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

>>>>

We performed a number of additional comparisons to evaluate the extent to which the thermal reactionomes differed between *A. carolinensis* and *A. picea*. First, we found that the profile of the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3) with a similar number of genes at high temperatures, more genes at intermediate temperatures in *A. carolinensis* and more genes at low temperatures in *A. picea*.

@@

We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis*. Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ... - ...%) was due to shifts in mean and ...% (95% CI ... - ...%) due to shifts in slope.

We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A. carolinensis* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`) and *Intermediate* genes in both (Wilcoxon V = `r w3$statistic`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r w3$conf.int[1:2]`), basal expression near optimum temperatures of 19.5C was always higher in *A. carolinensis*.

Sensitivity of thermally-responsive genes is indicated by the thermal range of *Bimodal* expressed genes.

We tested for a transcriptome-wide pattern of thermal sensitivity and tolerance by examining the thermal range for the *Bimodal* and *Intermediate* expressed genes. We found no in thermal sensitivity between species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76). However, for *Intermediate* genes, *A. carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
